---
title: "Practice plotting and reshaping data"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Unicef mortality rates
You are an employee at Unicef working to reduce neonatal mortality rates. 
You have compiled data looking at infant mortality under age 5. 
Rates are measured  as deaths per 1000 births for each year. 
```{r, message = FALSE}
library(tidyverse)
library(wesanderson) #just a fun color palette
mortality <- read.csv('~/Downloads/unicef-u5mr.csv',check.names=FALSE) #check.names:
  #deals with a naming issue because 
  #col names are numbers. 
```
\newpage
You began tracking these rates in 2000. 
Generate the plot below to compare mortality rates across countries in the year 2000. 
Try to generate the exact plot shown including the title and axis labels. 
Google is your friend! 
(tips: coord_flip() will filp your bar chart. The following notation lets you call column names that are numbers  ' 2000 '. )

&nbsp;
```{r, echo=FALSE}

```
\newpage
##comparing neonatal mortality over time. 
Based on the data shown in the first plot, you and your team decide to focus
interventions on three of the countries with the highest rates of neonatal mortality (Sierra Leone, Angola and Niger). 

Generate the following plot to show the change in mortality rates over time for just these three countries. 

(This may require some cleaning and reshaping). 

&nbsp;

```{r, echo = FALSE}

```
\newpage
See if you can find a way to customize the color of your plot.
tip: you can do this manually or with a palette (eg. brewer makes a color blind friendly palette)
I know my boss is a fan of wes anderson so in order to get their attention I'm going to use the fun wes anderson color palette. 


&nbsp;

```{r, echo=FALSE}

```
\newpage
Try splitting plots by country 
```{r, echo=FALSE}

```
\newpage
Think about what you learned last class. Come up with your own plot to visualize something interesting about this data. 
You can focus on a specific year, country or some comparison between them. 

Alternatively, use the plots we already generated but change the aesthetics (eg. the fill, colors, theme, lengend sizes). 

```{r, echo = FALSE}

```
